---
title: "Analysis of False Belief Task (2)"
author: "Sean Trott"
date: "April 24, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load LLM data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/open_llm_tom/src/analysis")
directory_path <- "../../data/processed/fb2_local/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_all_models <- bind_rows(csv_list) %>%
  mutate(model_shorthand = str_to_title(model_shorthand))
nrow(df_all_models)


length(unique(df_all_models$model_shorthand))
table(df_all_models$model_path)
table(df_all_models$model_shorthand)


### Load #params
df_model_properties = read_csv("../../data/processed/model_properties.csv") %>% 
  mutate(model_shorthand = str_to_title(model_shorthand))
nrow(df_model_properties)

### Load training data
df_training_data = read_csv("../../data/processed/model_training_data.csv")%>% 
  mutate(model_shorthand = str_to_title(model_shorthand))
table(df_training_data$model_shorthand)

### Join
df_all_properties = df_training_data %>%
  inner_join(df_model_properties)

table(df_all_properties$model_shorthand)
nrow(df_all_properties)


#### Bind with FB data
df_all_models = df_all_models %>%
  inner_join(df_all_properties)
nrow(df_all_models)
table(df_all_models$model_shorthand)

df_all_models = df_all_models %>%
  mutate(model_shorthand = str_to_title(model_shorthand))


length(unique(df_all_models$model_shorthand))
length(unique(df_all_models$model_family))


```


# LLM Analysis

For the main comparison to human data, we focus only on the `implicit` items, i.e., where knowledge states aren't directly stated.

## Sensitivity to mental states


Each model on its own:

```{r}
df_all_models %>%
  filter(knowledge_cue == "implicit") %>%
  mutate(condition = str_to_title(condition)) %>%
  filter(model_shorthand %in% c("Human",
                                "Olmo 2 32b Dpo",
                                "Olmo 2 32b Instruct",
                                "Olmo 2 32b Sft",
                                "Olmo 2 32b")) %>%
  ggplot(aes(x = log_odds,
             y = model_shorthand,
             fill = condition)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .8,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15))


# Function to fit models and perform model comparison
compare_models <- function(df) {
  if (n_distinct(df$condition) < 2) return(NULL)  # Skip if only one condition
  
  mod_full <- tryCatch(
    lmer(log_odds ~ condition +
           (1 | item), data = df, REML = FALSE),
    error = function(e) NULL
  )
  
  mod_reduced <- tryCatch(
      lmer(log_odds ~ 1 + (1 | item), data = df, REML = FALSE),
      error = function(e) NULL
  )
  
  if (is.null(mod_full) || is.null(mod_reduced)) return(NULL)
  
  anova_result <- anova(mod_full, mod_reduced)
  delta_aic <- AIC(mod_reduced) - AIC(mod_full)
  lrt_stat <- anova_result$Chisq[2]
  p_val <- anova_result$`Pr(>Chisq)`[2]
  
  # Extract coefficients
  coefs <- tryCatch(fixef(mod_full), error = function(e) return(rep(NA, 2)))
  cond_coef <- coefs[grep("^condition", names(coefs))]

  tibble(
    model_path = unique(df$model_path),
    delta_AIC = delta_aic,
    LRT_stat = lrt_stat,
    p_value = p_val,
    condition_coef = cond_coef
  )
}

# Apply to each model_path
results_by_model_path <- df_all_models %>%
  filter(knowledge_cue == "implicit") %>%
  group_by(model_path) %>%
  group_split() %>%
  map_dfr(compare_models)

results_by_model_path %>%
  arrange(LRT_stat)

summary(results_by_model_path$LRT_stat)
summary(results_by_model_path$condition_coef)


results_by_model_path$p_adj = p.adjust(results_by_model_path$p_value, method = "holm")
results_by_model_path = results_by_model_path %>%
  mutate(sig = p_adj < .05)
mean(results_by_model_path$sig)

results_by_model_path %>%
  filter(sig == TRUE) %>%
  select(model_path, LRT_stat, p_adj, condition_coef)

summary(results_by_model_path$condition_coef)
```


All models together:

```{r}

df_all_models_implicit = df_all_models %>%
  filter(knowledge_cue == "implicit")
### Mixed models
mod_full = lmer(data = df_all_models_implicit,
                log_odds ~ condition +
                  (1 + condition | model_path) + (1 + condition | item),
                REML = FALSE)

mod_reduced = lmer(data = df_all_models_implicit,
                log_odds ~ # knowledge_cue  + 
                  (1 + condition | model_path) + (1 + condition | item),
                REML = FALSE)

summary(mod_full)
anova(mod_full, mod_reduced)
```


## Accuracy metric

We first focus on the **implicit** items, since that's what we have human data for.

```{r}
df_all_models = df_all_models %>%
  mutate(correct = case_when(
    condition == "aware" & log_odds > 0 ~ TRUE,
    condition == "unaware" & log_odds <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  ))


df_summ = df_all_models %>%
  filter(knowledge_cue == "implicit") %>%
  group_by(model_path, model_shorthand, model_family,
           num_params, base_instruct, num_training_tokens) %>%
  summarise(mean_accuracy = mean(correct),
            n())

df_summ %>%
  select(model_shorthand, mean_accuracy) %>%
  arrange(-mean_accuracy) %>%
  head(5)
mean(df_all_models$correct)



df_summ %>%
  ungroup() %>%
  arrange(desc(mean_accuracy)) %>%
  select(model_shorthand, mean_accuracy) %>%
  head(5)


### "Wisdom of the crowd"?
df_lo_avg = df_all_models %>%
  group_by(passage, condition) %>%
  summarise(m_lo = mean(log_odds)) %>%
  mutate(correct = case_when(
    condition == "aware" & m_lo > 0 ~ TRUE,
    condition == "unaware" & m_lo <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  ))
mean(df_lo_avg$correct)




df_summ %>%
  ggplot(aes(x = num_params,
             y = mean_accuracy,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Parameters",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(8, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom")




### Which properties predict mean accuracy?
summary(lm(data = df_summ,
           mean_accuracy ~ log10(num_params) +
             base_instruct + log10(num_training_tokens)))




df_summ = df_all_models %>%
  filter(knowledge_cue == "implicit") %>%
  filter(item <= 8) %>%
  group_by(model_path, model_shorthand, model_family,
           num_params, base_instruct, num_training_tokens) %>%
  summarise(mean_accuracy = mean(correct),
            n())

df_summ %>%
  select(model_shorthand, mean_accuracy) %>%
  arrange(-mean_accuracy) %>%
  head(5)
mean(df_all_models$correct)


df_summ %>%
  ggplot(aes(x = num_params,
             y = mean_accuracy,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  geom_hline(yintercept = .82,
             linetype = "dotted", color = "red",
             size = 1.2, alpha = .8) + 
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Parameters",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(8, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom")


```



## Human analysis

```{r}
df_human = read_csv("../../data/processed/human/exp1_aggregated.csv") 

df_human_critical = df_human %>%
  filter(exp_group == "Knowledge") %>%
  mutate(condition = case_when(
    condition == "Speaker Aware" ~ "aware",
    condition == "Speaker Unaware" ~ "unaware"
  )) %>%
  mutate(item = stimNum) %>%
  mutate(correct = case_when(
    condition == "aware" & categorization == "Yes" ~ TRUE,
    condition == "unaware" & categorization == "No" ~ TRUE,
    condition == "unaware" & categorization == "Yes" ~ FALSE,
    condition == "aware" & categorization == "No" ~ FALSE
  )) %>%
  select(subject, rt, categorization, correct, item, condition, rc.mean, exp.inf.mean)
  

df_human_critical <- df_human_critical %>%
  mutate(answer_binary = ifelse(categorization == "Yes", 1, 0))


llm_by_item <- df_all_models %>%
  filter(knowledge_cue == "implicit") %>%
  select(item, condition, log_odds, model_path, model_shorthand)

df_merged <- df_human_critical %>%
  inner_join(llm_by_item, by = c("item", "condition"))

```

Now, fit a baselines model for each LLM.


```{r}
fit_compare_glmer_by_model <- function(df) {
  if (n_distinct(df$answer_binary) < 2) return(NULL)  # skip degenerate cases

  # Fit full model
  mod_full <- tryCatch(
    glmer(answer_binary ~ condition + log_odds +
            (1 | item) + (1|subject),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  # Fit reduced model (no condition)
  mod_reduced <- tryCatch(
    glmer(answer_binary ~ log_odds +
            (1 | item) + (1|subject),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  if (is.null(mod_full) || is.null(mod_reduced)) return(NULL)

  # Model comparison (Likelihood Ratio Test)
  anova_result <- anova(mod_reduced, mod_full)
  lrt_stat <- anova_result$Chisq[2]
  p_val <- anova_result$`Pr(>Chisq)`[2]
  delta_aic <- AIC(mod_reduced) - AIC(mod_full)

  # Extract coefficient for log_odds
  log_odds_coef <- tryCatch({
    fixef(mod_full)["log_odds"]
  }, error = function(e) NA)

  tibble(
    model_shorthand = unique(df$model_shorthand),
    delta_AIC = delta_aic,
    LRT_stat = lrt_stat,
    p_value = p_val,
    log_odds_coef = log_odds_coef,
    condition_coef = fixef(mod_full)["conditionunaware"],
    
  )
}



# Apply across model_shorthand groups
glmer_model_comparisons <- df_merged %>%
  group_by(model_shorthand) %>%
  group_split() %>%
  map_dfr(fit_compare_glmer_by_model)

glmer_model_comparisons

summary(glmer_model_comparisons$log_odds_coef)
summary(glmer_model_comparisons$condition_coef)
summary(glmer_model_comparisons$LRT_stat)

```


Compare effects:

```{r}

#### Do this for condition
# Calculate mean Log Odds for each model andcondition
df_all_models$answer = df_all_models$log_odds > 0

df_by_condition <- df_all_models %>%
  group_by(model_path, condition) %>%
  summarise(mean_yes = mean(answer),
            .groups = "drop")

# Calculate overall mean across all LMs
df_overall_mean <- df_all_models %>%
  group_by(condition) %>%
  summarise(mean_yes = mean(answer),
            .groups = "drop") %>%
  mutate(model_path = "All LMs Mean")

df_human_mean = df_human_critical %>%
  filter(rc.mean >= 9.5) %>%
  group_by(condition) %>%
  summarise(mean_yes = mean(answer_binary),
            .groups = "drop") %>%
  mutate(model_path = "Human")




# Add labels to the data
set.seed(1)
df_overall_mean <- df_overall_mean %>% mutate(label = "LLM Mean")
df_human_mean <- df_human_mean %>% mutate(label = "Human")
ggplot(df_by_condition, aes(x = condition, y = mean_yes)) +
  geom_jitter(width = 0.1, alpha = 0.3, size = 2, color = "gray40") +
  geom_point(data = df_overall_mean, 
             aes(x = condition, y = mean_yes),
             color = "#1f77b4", size = 5, shape = 18) +
  geom_text_repel(data = df_overall_mean,
                  aes(x = condition, y = mean_yes, label = label),
                  size = 4, color = "#1f77b4") +
  geom_point(data = df_human_mean, 
             aes(x = condition, y = mean_yes),
             color = "#d62728", size = 5, shape = 17) +
  geom_text_repel(data = df_human_mean,
                  aes(x = condition, y = mean_yes, label = label),
                  size = 4, color = "#d62728") +
  theme_minimal() +
  labs(x = "Knowledge State",
       y = "Proportion of Yes Responses") +
  theme(text = element_text(size = 15))



# Calculate human difference
df_human_diff <- df_human_mean %>%
  pivot_wider(names_from = condition, values_from = mean_yes) %>%
  mutate(diff = `aware` - `unaware`)

human_diff <- df_human_diff$diff
# Calculate the difference for each model
df_cond_diff <- df_by_condition %>%
  pivot_wider(names_from = condition, values_from = mean_yes) %>%
  mutate(diff = `aware` - `unaware`)  # Adjust these names to match your levels

# Calculate mean difference
mean_diff <- mean(df_cond_diff$diff)

# Plot the distribution of differences
ggplot(df_cond_diff, aes(x = diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, alpha = 0.4) +
  geom_density(color = "#1f77b4", linewidth = 1.2) +
  geom_vline(xintercept = mean_diff, color = "#1f77b4", linewidth = 1.5, linetype = "dashed") +
  geom_vline(xintercept = human_diff, color = "#d62728", linewidth = 1.5, linetype = "solid") +
  geom_vline(xintercept = 0, color = "black", linetype = "dotted") +
  annotate("text", x = mean_diff, y = Inf, label = "LLM Mean",
           hjust = -0.1, vjust = 1.5, color = "#1f77b4", size = 4) +
  annotate("text", x = human_diff, y = Inf, label = "Human",
           hjust = 1.1, vjust = 3, color = "#d62728", size = 4) +
  theme_minimal() +
  labs(x = "Difference in Proportion 'Yes' Responses (Aware - Unaware)",
       y = "Density") +
  theme(text = element_text(size = 15))

```


Restrict human accuracy by reading comprehension, etc.

```{r}
summary(df_human_critical$rc.mean)
summary(df_human_critical$exp.inf.mean)

df_human_critical %>% 
  filter(rc.mean >= 9.5) %>%
  summarise(mean_accuracy = mean(correct))

df_human_critical %>% 
  filter(exp.inf.mean >= 7.5) %>%
  summarise(mean_accuracy = mean(correct))




```


