---
title: "Analysis of False Belief Task"
author: "Sean Trott"
date: "April 24, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lmtest)
library(forcats)
library(broom)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```

# Load LLM data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/open_llm_tom/src/analysis")
directory_path <- "../../data/processed/fb_local/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_all_models <- bind_rows(csv_list) %>%
  mutate(model_shorthand = str_to_title(model_shorthand))
nrow(df_all_models)


length(unique(df_all_models$model_shorthand))
table(df_all_models$model_path)
table(df_all_models$model_shorthand)



### Load #params
df_model_properties = read_csv("../../data/processed/model_properties.csv") %>% 
  mutate(model_shorthand = str_to_title(model_shorthand))
nrow(df_model_properties)

### Load training data
df_training_data = read_csv("../../data/processed/model_training_data.csv")%>% 
  mutate(model_shorthand = str_to_title(model_shorthand))
table(df_training_data$model_shorthand)

### Join
df_all_properties = df_training_data %>%
  inner_join(df_model_properties)

table(df_all_properties$model_shorthand)
nrow(df_all_properties)


#### Bind with FB data
df_all_models = df_all_models %>%
  inner_join(df_all_properties)
nrow(df_all_models)
table(df_all_models$model_shorthand)

df_all_models = df_all_models %>%
  mutate(model_shorthand = str_to_title(model_shorthand))


length(unique(df_all_models$model_shorthand))
length(unique(df_all_models$model_family))



df_all_models <- df_all_models %>%
  mutate(model_shorthand = recode(model_shorthand,
    "Gemma 2 2b"          = "Gemma 1 2b",
    "Gemma 2 2b Instruct" = "Gemma 1 2b Instruct"
  ))

```


# LLM Analysis

## Sensitivity to mental states

```{r sensitivity}
### Visualization
df_all_models %>%
  ggplot(aes(x = log_odds,
             y = reorder(model_shorthand, num_params),
             fill = condition)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .8,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  facet_wrap(~knowledge_cue)

### Visualization by model family
df_all_models %>%
  group_by(passage, condition, knowledge_cue, model_family) %>%
  summarise(m_lo = mean(log_odds)) %>%
  ggplot(aes(x = m_lo,
             y = model_family,
             fill = condition)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .8,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  facet_wrap(~knowledge_cue)

### Visualization altogether
df_all_models %>%
  group_by(passage, condition, knowledge_cue) %>%
  summarise(m_lo = mean(log_odds)) %>%
  ggplot(aes(x = m_lo,
             y = knowledge_cue,
             fill = condition)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .8,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15))




```

Each model on its own:

```{r}
# Function to fit models and perform model comparison
compare_models <- function(df) {
  if (n_distinct(df$condition) < 2) return(NULL)  # Skip if only one condition
  
  mod_full <- tryCatch(
    lmer(log_odds ~ condition + knowledge_cue + first_mention + recent_mention +
           (1 + condition | start), data = df, REML = FALSE),
    error = function(e) NULL
  )
  
  mod_reduced <- tryCatch(
    lmer(log_odds ~ knowledge_cue + first_mention + recent_mention +
           (1 + condition | start), data = df, REML = FALSE),
    error = function(e) NULL
  )
  
  if (is.null(mod_full) || is.null(mod_reduced)) return(NULL)
  
  anova_result <- anova(mod_full, mod_reduced)
  delta_aic <- AIC(mod_reduced) - AIC(mod_full)
  lrt_stat <- anova_result$Chisq[2]
  p_val <- anova_result$`Pr(>Chisq)`[2]
  
  # Extract coefficients
  coefs <- tryCatch(fixef(mod_full), error = function(e) return(rep(NA, 2)))
  cond_coef <- coefs[grep("^condition", names(coefs))]
  cue_coef <- coefs[grep("^knowledge_cue", names(coefs))]

  tibble(
    model_path = unique(df$model_path),
    delta_AIC = delta_aic,
    LRT_stat = lrt_stat,
    p_value = p_val,
    condition_coef = cond_coef,
    knowledge_cue_coef = cue_coef
  )
}

# Apply to each model_path
results_by_model_path <- df_all_models %>%
  group_by(model_path) %>%
  group_split() %>%
  map_dfr(compare_models)
results_by_model_path %>%
  arrange(LRT_stat)

summary(results_by_model_path$LRT_stat)
summary(results_by_model_path$condition_coef)
summary(results_by_model_path$knowledge_cue_coef)


### TODO: Should we adjust? What hypothesis are we testing?
results_by_model_path$p_adj = p.adjust(results_by_model_path$p_value, method = "holm")
results_by_model_path = results_by_model_path %>%
  mutate(sig = p_adj < .05)
mean(results_by_model_path$sig)

results_by_model_path %>%
  filter(sig == TRUE) %>%
  select(model_path, LRT_stat, p_adj, condition_coef)
```


All models together:

```{r}

### Mixed models
mod_full = lmer(data = df_all_models,
                log_odds ~ condition + knowledge_cue + first_mention + recent_mention + 
                  (1 + condition | model_path) + (1 + condition | start),
                REML = FALSE)

mod_reduced = lmer(data = df_all_models,
                log_odds ~ knowledge_cue + first_mention + recent_mention + 
                  (1 + condition | model_path) + (1 + condition | start),
                REML = FALSE)

summary(mod_full)
anova(mod_full, mod_reduced)
```


## Accuracy metric


```{r accuracy}
df_all_models = df_all_models %>%
  mutate(correct = case_when(
    condition == "False Belief" & log_odds > 0 ~ TRUE,
    condition == "True Belief" & log_odds <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  ))



df_summ = df_all_models %>%
  group_by(model_path, model_shorthand, model_family,
           num_params, num_training_tokens, base_instruct) %>%
  summarise(mean_accuracy = mean(correct))
df_summ %>%
  select(model_shorthand, mean_accuracy)
mean(df_all_models$correct)

df_summ %>%
  ungroup() %>%
  arrange(desc(mean_accuracy)) %>%
  select(model_shorthand, mean_accuracy) %>%
  head(5)


### "Wisdom of the crowd"?
df_lo_avg = df_all_models %>%
  group_by(passage, condition) %>%
  summarise(m_lo = mean(log_odds)) %>%
  mutate(correct = case_when(
    condition == "False Belief" & m_lo > 0 ~ TRUE,
    condition == "True Belief" & m_lo <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  )) 
mean(df_lo_avg$correct)



df_summ %>%
  ggplot(aes(x = num_params,
             y = mean_accuracy,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  geom_hline(yintercept = .83,##TODO: Calculate from scratch
             linetype = "dotted", color = "red",
             size = 1.2, alpha = .8) + 
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "#Parameters (Log10)",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes

df_summ %>%
  ggplot(aes(x = num_training_tokens,
             y = mean_accuracy,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  geom_hline(yintercept = .83,##TODO: Calculate from scratch
             linetype = "dotted", color = "red",
             size = 1.2, alpha = .8) + 
  geom_hline(yintercept = .5, linetype = "dotted",
             size = 1.2, alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "#Training Tokens",
       y = "Accuracy",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes



### How do model properties predict the probability of a correct response?
mod_full = glmer(data = df_all_models,
                 correct ~ condition + knowledge_cue  +
                   log10(num_params) + log10(num_training_tokens) + 
                  base_instruct + first_mention + recent_mention +
                   (1 | start) + (1|model_family), 
                 family = binomial())
summary(mod_full)



### Plot coefficients
df_coef <- broom.mixed::tidy(mod_full, effects = "fixed") %>%
  mutate(term = forcats::fct_reorder(term, estimate))


df_coef %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error),
                width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    x = NULL, y = "Coefficient Estimate",
  ) +
  theme_minimal(base_size = 12)


write.csv(df_summ, "../../data/processed/summaries/fb_summary.csv")


### Using accuracy instead of binary correct/incorrect response

### Just individual predictors
summary(lm(data = df_summ,
           mean_accuracy ~ log10(num_params)))
summary(lm(data = df_summ,
           mean_accuracy ~ log10(num_training_tokens)))
summary(lm(data = df_summ,
           mean_accuracy ~ base_instruct))

### All together
summary(lm(data = df_summ,
           mean_accuracy ~ log10(num_params) + log10(num_training_tokens) + base_instruct))

```



# Comparison to human baseline

## Load human data

```{r}
df_human = read_csv("../../data/processed/human/human_fb_cleaned.csv") %>%
  select(participant_id, item_id, passage,
         is_correct, is_start, is_end,
         reaction_time, condition, response, first_mention, recent_mention, knowledge_cue)
mean(df_human$is_correct)

df_by_item = df_human %>%
  group_by(condition, passage, knowledge_cue) %>%
  summarise(p_start = mean(is_start) + .0001,
            p_end = mean(is_end) + .0001) %>%
  mutate(log_odds = log((p_start / p_end))) %>%
  mutate(model_shorthand = "Human") %>%
  mutate(correct = case_when(
    condition == "False Belief" & log_odds > 0 ~ TRUE,
    condition == "True Belief" & log_odds <= 0 ~ TRUE,
    TRUE ~ FALSE  # all other cases are incorrect
  ))
### Mean using log-odds measure forh umans
mean(df_by_item$correct)



### Merge
df_merged = df_all_models %>%
  bind_rows(df_by_item)
```


## Top-performing models

```{r best_models}
df_merged %>%
  filter(model_shorthand %in% c("Human",
                                "Olmo 2 13b Dpo",
                                "Olmo 2 13b Instruct",
                                "Olmo 2 13b Sft",
                                "Llama 3 70b Instruct",
                                "Olmo 2 13b")) %>%
  ggplot(aes(x = log_odds,
             y = model_shorthand,
             fill = condition)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color=NA, 
                       scale=.85, 
                       # size=1, 
                       alpha = .8,
                       stat="density") +
  labs(x = "Log Odds (Start vs. End)",
       y = "",
       fill = "") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme(
    legend.position = "bottom"
  ) + 
  theme(axis.title = element_text(size=rel(1.2)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text.x = element_text(size = rel(1.2))) +
    scale_fill_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  facet_wrap(~knowledge_cue) 

```



## Correlation matrix

```{r corr_matrix}
df_model_means = df_all_models %>%
  group_by(passage, condition, knowledge_cue) %>%
  summarise(log_odds = mean(log_odds)) %>%
  mutate(model_shorthand = "All Models")

df_wide <- df_merged %>%
  select(passage, model_shorthand, log_odds) %>%
  pivot_wider(names_from = model_shorthand, values_from = log_odds)

cor_matrix <- df_wide %>%
  select(-passage) %>%
  cor(use = "pairwise.complete.obs")

# Plot the correlation matrix
ggcorrplot(cor_matrix, 
           hc.order = FALSE,
           method = "square") +
  theme(
    axis.text.x = element_text(size = 6, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6)
  )

sort(cor_matrix["Human",])

### MDS - EXCLUDE HUMAN
# Remove Human from correlation matrix for MDS
cor_matrix_no_human <- cor_matrix[rownames(cor_matrix) != "Human", 
                                   colnames(cor_matrix) != "Human"]

# Convert correlation matrix to a dissimilarity matrix
dissimilarity <- as.dist(sqrt(1 - cor_matrix_no_human))

# Perform classical MDS
mds_result <- cmdscale(dissimilarity, k = 2)

# Put into a data frame for plotting
mds_df <- as.data.frame(mds_result)
colnames(mds_df) <- c("Dim1", "Dim2")
mds_df$model_shorthand <- rownames(mds_df)

mds_df = mds_df %>%
  inner_join(df_summ)

cor(mds_df$Dim1, mds_df$mean_accuracy)
cor(mds_df$Dim2, mds_df$mean_accuracy)

mds_df = mds_df %>%
  inner_join(results_by_model_path)

# Identify top-performing models for labeling
top_n_models <- 5  # Adjust this number
top_models <- mds_df %>%
  filter(model_shorthand %in% c("Qwen 2.5 7b Instruct", "Olmo 2 13b Dpo")) %>% 
  pull(model_shorthand)

# Add a column to indicate if model should be labeled
mds_df <- mds_df %>%
  mutate(label_model = ifelse(model_shorthand %in% top_models, 
                               model_shorthand, 
                               NA))

# Plot with labels
ggplot(mds_df, aes(Dim1, Dim2, 
                   color = model_family,
                   size = mean_accuracy,
                   shape = base_instruct)) +
  geom_point(alpha = .5) +
  geom_text(aes(label = label_model), 
            size = 3, 
            hjust = -0.1, 
            vjust = 0,
            show.legend = FALSE) +
  theme_minimal() +
  labs(x = "MDS 1",
       y = "MDS 2",
       color = "") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                   begin = 0.8, end = 0.15)) +
  guides(size = "none",      # Remove size legend
         shape = "none")




# Convert correlation matrix to long format
# Just extract the Human row/column directly
human_cors <- data.frame(
  model_shorthand = names(cor_matrix["Human", ]),
  correlation = cor_matrix["Human", ]
) %>%
  filter(model_shorthand != "Human") %>%
  arrange(desc(correlation))

rownames(human_cors) <- NULL

human_cors = human_cors %>%
  mutate(r2 = correlation ** 2)

summary(human_cors$r2)

human_cors %>%
  head(3)

human_cors = human_cors %>%
  inner_join(df_summ) 


human_cors %>%
  ggplot(aes(x = num_params,
             y = r2,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "#Parameters (Log10)",
       y = "R2",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes



mod_ppp = lm(data = human_cors,
               r2 ~ # mean_accuracy + 
                 log10(num_params) + log10(num_training_tokens)+ base_instruct)

summary(mod_ppp)
```


## Baselines analysis

First, merge with raw human data.

```{r}
df_human_shortened = df_human %>%
  # mutate(human_is_start = is_start) %>%
  select(participant_id, item_id, passage, condition, knowledge_cue, 
         is_start, reaction_time)

df_all_models_with_human = df_all_models %>%
  inner_join(df_human_shortened)

table(df_all_models_with_human$model_path)

```

Now, fit a baselines model for each LLM.


```{r}
fit_compare_glmer_by_model <- function(df) {
  if (n_distinct(df$is_start) < 2) return(NULL)  # skip degenerate cases

  # Fit full model
  mod_full <- tryCatch(
    glmer(is_start ~ condition + knowledge_cue + 
            first_mention + recent_mention + log_odds +
            (1 | item_id),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  # Fit reduced model (no condition)
  mod_reduced <- tryCatch(
    glmer(is_start ~ knowledge_cue + 
            first_mention + recent_mention + log_odds +
            (1 | item_id),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  if (is.null(mod_full) || is.null(mod_reduced)) return(NULL)

  # Model comparison (Likelihood Ratio Test)
  anova_result <- anova(mod_reduced, mod_full)
  lrt_stat <- anova_result$Chisq[2]
  p_val <- anova_result$`Pr(>Chisq)`[2]
  delta_aic <- AIC(mod_reduced) - AIC(mod_full)

  # Extract coefficient for log_odds
  log_odds_coef <- tryCatch({
    fixef(mod_full)["log_odds"]
  }, error = function(e) NA)

  tibble(
    model_shorthand = unique(df$model_shorthand),
    delta_AIC = delta_aic,
    LRT_stat = lrt_stat,
    p_value = p_val,
    log_odds_coef = log_odds_coef,
    knowledge_cue_coef = fixef(mod_full)["knowledge_cueImplicit"],
    condition_coef = fixef(mod_full)["conditionTrue Belief"],
    
  )
}

# Apply across model_shorthand groups
glmer_model_comparisons <- df_all_models_with_human %>%
  group_by(model_shorthand) %>%
  group_split() %>%
  map_dfr(fit_compare_glmer_by_model)

glmer_model_comparisons

summary(glmer_model_comparisons$log_odds_coef)
summary(glmer_model_comparisons$knowledge_cue_coef)
summary(glmer_model_comparisons$condition_coef)
summary(glmer_model_comparisons$LRT_stat)

```


## PPP Analysis

```{r ppp}
fit_ppp_by_model <- function(df) {

  # Fit full model
  mod_full <- tryCatch(
    glmer(is_start ~ log_odds+first_mention + recent_mention + condition +
            (1 | item_id),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  # Extract coefficient for log_odds
  log_odds_coef <- tryCatch({
    fixef(mod_full)["log_odds"]
  }, error = function(e) NA)

  tibble(
    model_shorthand = unique(df$model_shorthand),
    AIC = AIC(mod_full),
    log_odds_coef = fixef(mod_full)["log_odds"]
  )
}

# Apply across model_shorthand groups
ppp_models <- df_all_models_with_human %>%
  group_by(model_shorthand) %>%
  group_split() %>%
  map_dfr(fit_ppp_by_model) %>%
  inner_join(df_model_properties)

mod_base <- glmer(is_start ~ first_mention + recent_mention + condition + # log_odds+
          (1 | item_id),
        data = filter(df_all_models_with_human, model_shorthand == "Qwen 2.5 7b Instruct"),
        family = binomial())
  
### Rescale AIC
ppp_models = ppp_models %>%
  mutate(delta_AIC = AIC - AIC(mod_base))

summary(ppp_models$delta_AIC)

ppp_models %>%
  arrange(delta_AIC) %>%
  head(5)


### Add back to original data
ppp_with_accuracy = ppp_models %>%
  select(model_shorthand, delta_AIC) %>%
  inner_join(df_summ)



ppp_with_accuracy %>%
  ggplot(aes(x = num_training_tokens,
             y = delta_AIC,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "#Training Tokens",
       y = "Delta AIC",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes


ppp_with_accuracy %>%
  ggplot(aes(x = num_params,
             y = delta_AIC,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "#Parameters",
       y = "Delta AIC",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes

### Among models that >50% accuracy, better models explain more variance in human data
ppp_with_accuracy %>%
  ggplot(aes(x = mean_accuracy,
             y = delta_AIC,
             color = model_family,
             shape = base_instruct)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Accuracy",
       y = "Delta AIC",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes



mod_ppp = lm(data = ppp_with_accuracy,
               delta_AIC ~ 
                 log10(num_params) + log10(num_training_tokens)+ base_instruct )

summary(mod_ppp)
```


## Knowledge cue

```{r kc_epistemics}
mod_full = glmer(is_start ~ condition + knowledge_cue + 
          first_mention + recent_mention +
          (1 + knowledge_cue | item_id),
        data = df_human,
        family = binomial())
mod_no_kc = glmer(is_start ~ condition + # knowledge_cue + 
          first_mention + recent_mention +
          (1+ knowledge_cue  | item_id),
        data = df_human,
        family = binomial())

anova(mod_full, mod_no_kc)
summary(mod_full)



### Comopare to parameter estimate for LLMs, using binary outcome
df_all_models$is_start = df_all_models$log_odds > 0


coef_human <- fixef(mod_full)

# Create aligned dataframe
group_coefs <- tibble(
  model_path = c("Human"),
  knowledge_cue_coef = c(coef_human["knowledge_cueImplicit"]),
  condition_coef = c(coef_human["conditionTrue Belief"])
)

## Each model on its own
run_models_for_comparison <- function(df) {

  mod_full = glmer(is_start ~ condition + knowledge_cue + 
            first_mention + recent_mention +
            (1 + condition | start),
          data = df,
          family = binomial(),
          control = glmerControl(optimizer = "bobyqa"))
  
  # Extract coefficients
  coefs <- fixef(mod_full)
  cond_coef <- coefs[grep("^condition", names(coefs))]
  cue_coef <- coefs[grep("^knowledge_cue", names(coefs))]

  tibble(
    model_path = unique(df$model_path),
    condition_coef = cond_coef,
    knowledge_cue_coef = cue_coef
  )
}

# Apply to each model_path
results_by_model_path <- df_all_models %>%
  group_by(model_path) %>%
  group_split() %>%
  map_dfr(run_models_for_comparison)


all_coefs = results_by_model_path %>%
  select(model_path, knowledge_cue_coef, condition_coef) 

summary(all_coefs$knowledge_cue_coef)
sd(all_coefs$knowledge_cue_coef)
sd(all_coefs$knowledge_cue_coef)/sqrt(nrow(all_coefs))

summary(all_coefs$condition_coef)
sd(all_coefs$condition_coef)
sd(all_coefs$condition_coef)/sqrt(nrow(all_coefs))

### Percentile of human effect (KC)
percentile <- mean(all_coefs$knowledge_cue_coef >= -1.02) * 100
percentile


### Percentile of human effect (condition)
percentile <- mean(all_coefs$condition_coef >= -3.78) * 100
percentile




### Merge with other data
df_all_coefs = all_coefs %>%
  inner_join(df_summ)

summary(lm(data = df_all_coefs, knowledge_cue_coef ~ log10(num_params) +
             log10(num_training_tokens) + base_instruct))


### Larger models show bigger effect
top_models <- df_all_coefs %>%
  filter(model_shorthand %in% c("Olmo 2 13b Dpo", "Olmo 2 13b")) %>% 
  pull(model_shorthand)

# Add a column to indicate if model should be labeled
df_all_coefs <- df_all_coefs %>%
  mutate(label_model = ifelse(model_shorthand %in% top_models, 
                               model_shorthand, 
                               NA))

df_all_coefs %>%
  ggplot(aes(x = num_params,
             y = knowledge_cue_coef,
             color = model_family,
             shape = base_instruct)) +
  geom_hline(yintercept = -1.02, color = "red",
             linetype = "dotted", size = 1.2, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = label_model), 
            size = 3, 
            hjust = -0.1, 
            vjust = 0,
            show.legend = FALSE) +
  theme_minimal() +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "#Parameters (Log10)",
       y = "Parameter Estimate (Knowledge Cue)",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes




df_all_coefs %>%
  ggplot(aes(x = num_params,
             y = condition_coef,
             color = model_family,
             shape = base_instruct)) +
  geom_hline(yintercept = -3.78, color = "red",
             linetype = "dotted", size = 1.2, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = label_model), 
            size = 3, 
            hjust = -0.1, 
            vjust = 0,
            show.legend = FALSE) +
  theme_minimal() +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  # geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "#Parameters (Log10)",
       y = "Parameter Estimate (Knowledge State)",
       color = "",
       shape = "") +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(5, option = "mako", 
                                                  begin = 0.8, end = 0.15)) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  guides(color = guide_legend(nrow = 2),     # Wrap colors
         shape = guide_legend(nrow = 2))     # Wrap shapes



df_all_coefs %>%
  ggplot(aes(x = knowledge_cue_coef,
             y = condition_coef,
             size = num_params)) +
  geom_point(alpha = .5) +
  geom_point(aes(x = -1.02, y = -3.78), color = "red") +
  annotate("text", x = -1.02, y = -3.78, label = "Human Estimate",
           hjust = .3, vjust = 1.5, color = "red", size = 4) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal() +
  labs(x = "Parameter Estimate (Knowledge Cue)",
       y = "Parameter Estimate (Knowledge State)") + 
  theme(text = element_text(size = 15),
        legend.position="none") 



#### Plot effect of KC across LMs
# Calculate mean Log Odds for each model and Knowledge Cue
df_by_kc <- df_all_models %>%
  group_by(model_path, knowledge_cue) %>%
  summarise(mean_start = mean(is_start),
            .groups = "drop")

# Calculate overall mean across all LMs
df_overall_mean <- df_all_models %>%
  group_by(knowledge_cue) %>%
  summarise(mean_start = mean(is_start),
            .groups = "drop") %>%
  mutate(model_path = "All LMs Mean")

df_human_mean = df_human %>%
  group_by(knowledge_cue) %>%
  summarise(mean_start = mean(is_start),
            .groups = "drop") %>%
  mutate(model_path = "Human")

# Plot
library(ggrepel)


# Add labels to the data
set.seed(1)
df_overall_mean <- df_overall_mean %>% mutate(label = "LLM Mean")
df_human_mean <- df_human_mean %>% mutate(label = "Human")
ggplot(df_by_kc, aes(x = knowledge_cue, y = mean_start)) +
  geom_jitter(width = 0.1, alpha = 0.3, size = 2, color = "gray40") +
  geom_point(data = df_overall_mean, 
             aes(x = knowledge_cue, y = mean_start),
             color = "#1f77b4", size = 5, shape = 18) +
  geom_text_repel(data = df_overall_mean,
                  aes(x = knowledge_cue, y = mean_start, label = label),
                  size = 4, color = "#1f77b4") +
  geom_point(data = df_human_mean, 
             aes(x = knowledge_cue, y = mean_start),
             color = "#d62728", size = 5, shape = 17) +
  geom_text_repel(data = df_human_mean,
                  aes(x = knowledge_cue, y = mean_start, label = label),
                  size = 4, color = "#d62728") +
  theme_minimal() +
  labs(x = "Knowledge Cue",
       y = "Proportion of Start Responses") +
  theme(text = element_text(size = 15))



# Calculate human difference
df_human_diff <- df_human_mean %>%
  pivot_wider(names_from = knowledge_cue, values_from = mean_start) %>%
  mutate(diff = Explicit - Implicit)

human_diff <- df_human_diff$diff
# Calculate the difference for each model
df_kc_diff <- df_by_kc %>%
  pivot_wider(names_from = knowledge_cue, values_from = mean_start) %>%
  mutate(diff = Explicit - Implicit)  # Adjust these names to match your levels

# Calculate mean difference
mean_diff <- mean(df_kc_diff$diff)

# Plot the distribution of differences
ggplot(df_kc_diff, aes(x = diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, alpha = 0.4) +
  geom_density(color = "#1f77b4", linewidth = 1.2) +
  geom_vline(xintercept = mean_diff, color = "#1f77b4", linewidth = 1.5, linetype = "dashed") +
  geom_vline(xintercept = human_diff, color = "#d62728", linewidth = 1.5, linetype = "solid") +
  geom_vline(xintercept = 0, color = "black", linetype = "dotted") +
  annotate("text", x = mean_diff, y = Inf, label = "LLM Mean",
           hjust = -0.1, vjust = 1.5, color = "#1f77b4", size = 4) +
  annotate("text", x = human_diff, y = Inf, label = "Human",
           hjust = -0.1, vjust = 3, color = "#d62728", size = 4) +
  theme_minimal() +
  labs(x = "Difference in Proportion Start (Explicit - Implicit)",
       y = "Density") +
  theme(text = element_text(size = 15))




#### Do this for condition
# Calculate mean Log Odds for each model andcondition
df_by_condition <- df_all_models %>%
  group_by(model_path, condition) %>%
  summarise(mean_start = mean(is_start),
            .groups = "drop")

# Calculate overall mean across all LMs
df_overall_mean <- df_all_models %>%
  group_by(condition) %>%
  summarise(mean_start = mean(is_start),
            .groups = "drop") %>%
  mutate(model_path = "All LMs Mean")

df_human_mean = df_human %>%
  group_by(condition) %>%
  summarise(mean_start = mean(is_start),
            .groups = "drop") %>%
  mutate(model_path = "Human")

# Plot
library(ggrepel)


# Add labels to the data
set.seed(1)
df_overall_mean <- df_overall_mean %>% mutate(label = "LLM Mean")
df_human_mean <- df_human_mean %>% mutate(label = "Human")
ggplot(df_by_condition, aes(x = condition, y = mean_start)) +
  geom_jitter(width = 0.1, alpha = 0.3, size = 2, color = "gray40") +
  geom_point(data = df_overall_mean, 
             aes(x = condition, y = mean_start),
             color = "#1f77b4", size = 5, shape = 18) +
  geom_text_repel(data = df_overall_mean,
                  aes(x = condition, y = mean_start, label = label),
                  size = 4, color = "#1f77b4") +
  geom_point(data = df_human_mean, 
             aes(x = condition, y = mean_start),
             color = "#d62728", size = 5, shape = 17) +
  geom_text_repel(data = df_human_mean,
                  aes(x = condition, y = mean_start, label = label),
                  size = 4, color = "#d62728") +
  theme_minimal() +
  labs(x = "Knowledge State",
       y = "Proportion of Start Responses") +
  theme(text = element_text(size = 15))



# Calculate human difference
df_human_diff <- df_human_mean %>%
  pivot_wider(names_from = condition, values_from = mean_start) %>%
  mutate(diff = `False Belief` - `True Belief`)

human_diff <- df_human_diff$diff
# Calculate the difference for each model
df_cond_diff <- df_by_condition %>%
  pivot_wider(names_from = condition, values_from = mean_start) %>%
  mutate(diff = `False Belief` - `True Belief`)  # Adjust these names to match your levels

# Calculate mean difference
mean_diff <- mean(df_cond_diff$diff)

# Plot the distribution of differences
ggplot(df_cond_diff, aes(x = diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, alpha = 0.4) +
  geom_density(color = "#1f77b4", linewidth = 1.2) +
  geom_vline(xintercept = mean_diff, color = "#1f77b4", linewidth = 1.5, linetype = "dashed") +
  geom_vline(xintercept = human_diff, color = "#d62728", linewidth = 1.5, linetype = "solid") +
  geom_vline(xintercept = 0, color = "black", linetype = "dotted") +
  annotate("text", x = mean_diff, y = Inf, label = "LLM Mean",
           hjust = -0.1, vjust = 1.5, color = "#1f77b4", size = 4) +
  annotate("text", x = human_diff, y = Inf, label = "Human",
           hjust = 1.1, vjust = 3, color = "#d62728", size = 4) +
  theme_minimal() +
  labs(x = "Difference in Proportion Start (FB - TB)",
       y = "Density") +
  theme(text = element_text(size = 15))


```

### Baselines analysis (knowledge cue)

Now, fit a baselines model for each LLM.


```{r}
fit_compare_glmer_by_model <- function(df) {
  if (n_distinct(df$is_start) < 2) return(NULL)  # skip degenerate cases

  # Fit full model
  mod_full <- tryCatch(
    glmer(is_start ~ condition + knowledge_cue + 
            first_mention + recent_mention + log_odds +
            (1 | item_id),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  # Fit reduced model (no condition)
  mod_reduced <- tryCatch(
    glmer(is_start ~ condition + 
            first_mention + recent_mention + log_odds +
            (1 | item_id),
          data = df,
          family = binomial()),
    error = function(e) NULL
  )

  if (is.null(mod_full) || is.null(mod_reduced)) return(NULL)

  # Model comparison (Likelihood Ratio Test)
  anova_result <- anova(mod_reduced, mod_full)
  lrt_stat <- anova_result$Chisq[2]
  p_val <- anova_result$`Pr(>Chisq)`[2]
  delta_aic <- AIC(mod_reduced) - AIC(mod_full)

  # Extract coefficient for log_odds
  log_odds_coef <- tryCatch({
    fixef(mod_full)["log_odds"]
  }, error = function(e) NA)

  tibble(
    model_shorthand = unique(df$model_shorthand),
    delta_AIC = delta_aic,
    LRT_stat = lrt_stat,
    p_value = p_val,
    log_odds_coef = log_odds_coef,
    knowledge_cue_coef = fixef(mod_full)["knowledge_cueImplicit"],
    condition_coef = fixef(mod_full)["conditionTrue Belief"],
    
  )
}

# Apply across model_shorthand groups
glmer_model_comparisons <- df_all_models_with_human %>%
  group_by(model_shorthand) %>%
  group_split() %>%
  map_dfr(fit_compare_glmer_by_model)

glmer_model_comparisons

summary(glmer_model_comparisons$log_odds_coef)
summary(glmer_model_comparisons$knowledge_cue_coef)
summary(glmer_model_comparisons$condition_coef)
summary(glmer_model_comparisons$LRT_stat)

```

